\documentclass[smallextended]{svjour3}
\usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathrsfs}
\usepackage{url}
\usepackage{eucal}

\journalname{Semigroup Forum}

\def\om{\omega}
\def\Rc{\mathrel{\mathscr{R}}}
\def\Lc{\mathrel{\mathscr{L}}}

\newcommand{\sgp}{semi\-group}
\newcommand{\sgps}{semi\-groups}
\newcommand{\is}{involutory semi\-group}
\newcommand{\iss}{involutory semi\-groups}
\newcommand{\fis}{finite involutory semi\-group}
\newcommand{\fiss}{finite involutory semi\-groups}
\newcommand{\fs}{finite \sgp}
\newcommand{\fss}{finite \sgps}
\newcommand{\fus}{finite unary \sgp}
\newcommand{\fuss}{finite unary \sgps}
\newcommand{\fb}{finitely based}
\newcommand{\fbp}{finite basis problem}
\newcommand{\fg}{finitely generated}
\newcommand{\nfb}{non\-finitely based}
\newcommand{\infb}{inherently non\-finitely based}
\newcommand{\TSL}{\ensuremath{\mathcal{T\kern-.5pt S\kern-.5pt L}}}
\newcommand{\TB}{\ensuremath{\mathcal{T\kern-.5pt B}_2^1}}
\newcommand{\FI}{\ensuremath{\mathcal{F\kern-.3pt I}}}


\newtheorem{Thm}{Theorem}[section]
\newtheorem{Prop}[Thm]{Proposition}
\newtheorem{Lemma}[Thm]{Lemma}
\newtheorem{Def}{Definition}[section]
\newtheorem{Res}[Thm]{Result}
\newtheorem{Cor}[Thm]{Corollary}
\newtheorem{Example}[Thm]{Example}
\newtheorem{Examples}[Thm]{Examples}
\newtheorem{Conjecture}[Thm]{Conjecture}
\newtheorem{Not}[Thm]{Notation}

\renewcommand{\theThm}{\thesection.\arabic{Thm}}

\DeclareMathOperator{\var}{\mathsf{var}}

\begin{document}

\title{Unary enhancements\\ of inherently non-finitely based semigroups}
\author{K.~Auinger \and I.~Dolinka \and T.~V.~Pervukhina \and M.~V.~Volkov}

\institute{K.~Auinger \at Fakult\"at f\"ur Mathematik, Universit\"at Wien, Nordbergstrasse 15,  A-1090 Wien, Austria\\
\email{karl.auinger@univie.ac.at} \and
I.~Dolinka \at Department of Mathematics and Informatics, University of Novi Sad, Trg Dositeja Obradovi\'ca 4, 21101 Novi Sad, Serbia\\
\email{dockie@dmi.uns.ac.rs} \and
T.~V.~Pervukhina \at Institute of Mathematics and Computer Science, Ural Federal University, Lenina 51, 620000 Ekaterinburg, Russia\\
\email{tsungirl@yandex.ru} \and
M.~V.~Volkov \at Institute of Mathematics and Computer Science, Ural Federal University, Lenina 51, 620000 Ekaterinburg, Russia\\
\email{mikhail.volkov@usu.ru}}


\date {Received: / Accepted}

\maketitle

\begin{abstract}
This paper is a supplement to~\cite{ADV:2012}, more precisely, to Subsection 2.2 of~\cite{ADV:2012} dealing with \infb\ \iss. We exhibit a
simple condition under which a finite \is\ whose \sgp\ reduct is \infb\ is also \infb\ as a unary \sgp. As applications, we get both
already known and new examples of \infb\ \iss. We also show that for finite regular semigroups, our condition is not only sufficient but
also necessary for the property of being \infb\ to persist. This leads to an algorithmic description of regular \infb\ \iss.

\keywords{\is\ \and \infb\ \sgp\ \and twisted semilattice \and twisted Brandt monoid \and regular semigroup \and matrix semigroup}
\subclass{20M07 \and 20M17 \and 20M25}
\end{abstract}

\section{Background and overview}

The \fbp, that is the problem of classifying \sgps\ according to the finite basability of their identities, has been intensively explored
since the 1960s. Since the 1970s, the same problem has become investigated for \sgps\ endowed with an additional unary operation $x\mapsto
x^*$; such structures are commonly called \emph{unary \sgps}.

If $\mathcal{S}=\langle S,\cdot,{}^*\rangle$ is a unary \sgp, then the (plain) \sgp\ $\langle S,\cdot\rangle$ is called the (\emph{\sgp})
\emph{reduct} of $\mathcal{S}$. It is quite natural to ask how the answer to the \fbp\ for a given unary \sgp\ relates to the finite
basability of the identities of its reduct. The question turns out to be somewhat delicate. On the one hand, when we enhance the vocabulary
of the equational language by adding a unary operation, the expressive power of the language increases. Hence $\mathcal{S}$ usually has
more identities than $\langle S,\cdot\rangle$ so that the former may have more chance to become \nfb. On the other hand, the inference
power of the language increases too. Hence one can imagine the situation when some identity of $\langle S,\cdot\rangle$  does not follow
from an identity system $\Sigma$ as a plain identity but follows from $\Sigma$ when treated as a unary identity. This indicates that
$\mathcal{S}$ may be \fb\ even if $\langle S,\cdot\rangle$ is not. The cumulative effect of the trade-off between increased expressivity
and increased inference power is hard to predict in general, and both possible outcomes indeed occur. This means that there exist unary
\sgps, even groups $\mathcal{G}=\langle G;\cdot,{}^{-1}\rangle$ with inversion as the unary operation, such that $\mathcal{G}$ is \fb\
[\nfb] as a group while its reduct $\langle G;\cdot\rangle$ is \nfb\ [respectively, \fb] as a plain \sgp. See \cite[Section~2]{Volkov:2001}
for concrete examples (known since the 1970s), references and a discussion.

Much attention has been paid to the restriction of the \fbp\ to the class of \fss, in both plain and unary settings, see, e.g., the survey
\cite{Volkov:2001}. Therefore it appears a bit surprising that the above question about the relation between the finite basability of a
unary \sgp\ and of its reduct has not been systematically explored in the realm of \fss. To the best of our knowledge, the first example of
a \nfb\ \fus\ whose reduct is \fb\ was constructed only in~1998, see~\cite{Lawrence&Willard:1998}. The unary operation used
in~\cite{Lawrence&Willard:1998} was rather ad hoc, and similar examples with well behaved unary operations (including an example of a \nfb\
\fis\ with \fb\ reduct) have only recently appeared in \cite{Jackson&Volkov:2010}. The `opposite' examples (of \fb\ \fuss\ with \nfb\
reducts) are not yet known.

For \fss, the following strengthening of the property of being \nfb\ have been successfully studied. Recall that a variety $\mathbf{V}$ of
[unary] \sgps\ is called \emph{locally finite} if every \fg\ [unary] \sgp\ in $\mathbf{V}$ is finite. A finite [unary] \sgp\ is said to be
\emph{\infb} (INFB for short) if it is not contained in any locally finite \fb\ variety of [unary] \sgps. Since the variety generated by a
finite [unary] \sgp\ is known to be locally finite, an INFB [unary] \sgp\ certainly is \nfb. In fact, the property of being INFB is much
stronger than the property of being \nfb\ and also behaves more regularly, see \cite{Volkov:2001} for details.

Sapir~\cite{Sapir:1987} has given an efficient (in the algorithmic sense of the word) description of INFB \sgps. INFB unary \sgps\ have
been investigated in~\cite{Dolinka:2010,ADV:2012} where some sufficient conditions for a \fis\ to be INFB have been found. Again, in this
situation it is quite natural to ask what happens when one passes from a \fus\ to its reduct. The aforementioned example
of~\cite{Lawrence&Willard:1998} is in fact INFB so that in general an INFB unary \sgp\ may have a \fb\ reduct. This is however impossible
for a \fis; moreover, it is easy to verify (see Lemma~\ref{easy} below) that the reduct of an INFB \is\ must be INFB. The converse is not
true as first observed in~\cite{Sapir:1993}, and it is this circumstance that gives rise to the specific question addressed in the present
paper: when does an involution $x\mapsto x^*$ defined on an INFB \sgp\ $\langle S,\cdot\rangle$ preserve the property of being INFB in the
sense that the resulting \is\ $\mathcal{S}=\langle S,\cdot,{}^*\rangle$ is INFB as a unary \sgp?  We show (Theorem~\ref{twisted}) that this
is the case whenever the variety generated by $\mathcal{S}$ contains a certain 3-element \is\ $\TSL$ (twisted semilattice). This result has
several applications: first, we give new, easy and uniform proofs for some examples of INFB \iss\ found in~\cite{ADV:2012}; second, we
exhibit a further series of INFB \iss. We also show (Theorem~\ref{regular}) that if $\langle S,\cdot\rangle$ is a finite regular \sgp, then
the presence of the 3-element twisted semilattice $\TSL$ in the variety generated by $\mathcal{S}$ is not only sufficient but also
necessary for the property of being INFB to persist. Combined with Sapir's result, this leads to an efficient description of regular INFB
\iss.

\section{Preliminaries}
\label{sec:preliminaries}

We assume the reader's acquaintance with basic concepts of universal algebra such as the notion of a variety and the HSP-theorem, see,
e.g., \cite[Chapter~II]{BuSa81}. Section~\ref{sec:regular} also requires some knowledge of Green's relations, cf.~\cite[Chapter~2]{how}.

A unary \sgp\ $\mathcal{S}=\langle S,\cdot,{}^*\rangle$ is called an \emph{\is} if it satisfies the identities
\begin{equation}
\label{eq:invlution} (xy)^*=y^*x^* \quad \text{ and  }\quad (x^*)^*=x,
\end{equation}
in other words, if the unary operation $x\mapsto x^*$ is an involutory anti-automorphism of the reduct $\langle S,\cdot\rangle$.

The \emph{free \is} $\FI(X)$ on a given alphabet $X$ can be constructed as follows.  Let $\overline{X}=\{x^*\mid x\in X\}$ be a disjoint
copy of $X$. We refer to the elements of $X$ as \emph{plain letters} and to the elements of $\overline{X}$ as \emph{starred letters}.
Define $(x^*)^*=x$ for all $x^*\in \overline{X}$. Then $\FI(X)$ is the free \sgp\ $(X\cup\overline{X})^+$ endowed with the involution
defined by
$$(x_1\cdots x_m)^* = x_m^*\cdots x_1^*$$
for all $x_1,\dots,x_m\in X\cup \overline{X}$. We refer to elements of $\FI(X)$ as \emph{involutory words over $X$} while elements of the
free semigroup $X^+$ will be referred to as \emph{plain words over $X$}.

If an \is\ $\mathcal{T}=\langle T,\cdot,{}^*\rangle$ is generated by a set $Y\subseteq T$, then every element in $\mathcal{T}$ can be
represented by an involutory word over $Y$ and thus by a plain word over $Y\cup\overline{Y}$ where $\overline{Y}=\{y^*\mid y\in Y\}$. Hence
the reduct $\langle T,\cdot\rangle$ is generated by the set $Y\cup\overline{Y}$; in particular, $\mathcal{T}$ is \fg\ if and only if so is
$\langle T,\cdot\rangle$. This observation immediately leads to the following fact already mentioned in the introduction.

\begin{Lemma}
\label{easy} If an \is\ $\mathcal{S}=\langle S,\cdot,{}^*\rangle$ is \infb, then so is its reduct $\langle S,\cdot\rangle$.
\end{Lemma}

\begin{proof}
Arguing by contradiction, assume that $\langle S,\cdot\rangle$ is not INFB. Then $\langle S,\cdot\rangle$ belongs to a locally finite plain
\sgp\ variety defined by a finite identity system $\Sigma$. Consider the variety $\mathbf{V}$ of \iss\ defined by the identities
\eqref{eq:invlution} and $\Sigma$. Clearly, $\mathbf{V}$ is \fb\ and $\mathcal{S}\in\mathbf{V}$. If $\mathcal{T}=\langle
T,\cdot,{}^*\rangle$ is a \fg\ \is\ from $\mathbf{V}$, the reduct $\langle T,\cdot\rangle$ is a \fg\ plain \sgp\ by the observation
preceding the formulation of the lemma. Since the reduct satisfies the identities in $\Sigma$ and $\Sigma$ defines a locally finite plain
\sgp\ variety, we conclude that the base set $T$ is finite. Hence the variety $\mathbf{V}$ is locally finite and $\mathcal{S}$ belongs to a
locally finite \fb\ variety, a contradiction.
\end{proof}

As mentioned, the converse of Lemma~\ref{easy} is not true in general. For an example, consider the well known \emph{Brandt monoid}
$\langle B_2^1,\cdot\rangle$, where $B_2^1$ is the set of the following six integer $2\times 2$-matrices:
$$\begin{pmatrix} 0 & 0\\ 0 & 0\end{pmatrix},\
\begin{pmatrix} 1 & 0\\ 0 & 0\end{pmatrix},\
\begin{pmatrix} 0 & 1\\ 0 & 0\end{pmatrix},\
\begin{pmatrix} 0 & 0\\ 1 & 0\end{pmatrix},\
\begin{pmatrix} 0 & 0\\ 0 & 1\end{pmatrix},\
\begin{pmatrix} 1 & 0\\ 0 & 1\end{pmatrix},$$
and the binary operation $(A_1,A_2)\mapsto A_1\cdot A_2$ is the usual matrix multiplication. It is known
\cite[Corollary~6.1]{sapirburnside} that the Brandt monoid is INFB (this was in fact the very first example of an INFB \sgp). The Brandt
monoid admits a natural involution, namely, the usual matrix transposition $A\mapsto A^T$. However, the \is\ $\langle
B_2^1,\cdot,{}^T\rangle$ is not INFB as shown in~\cite{Sapir:1993}. Further examples can be found in~\cite{ADV:2012}: if $\mathcal{K}$ is a
finite field and $\mathrm{M}_n(\mathcal{K})$ stands for the set of all $n\times n$-matrices over $\mathcal{K}$, then the \sgp\
$\langle\mathrm{M}_n(\mathcal{K}),\cdot\rangle$ is INFB for any $n\ge 2$ by \cite[Corollary~6.2]{sapirburnside} while the \is\
$\langle\mathrm{M}_2(\mathcal{K}),\cdot,{}^T\rangle$ is not INFB if the number of elements in $\mathcal{K}$ leaves reminder~3 when divided
by~4.

Thus, not every involution defined on an INFB \sgp\ preserves the property of being INFB, and we are looking towards a classification of
`INFB-preserving' involutions. Our main tool is the following result, which is Theorem 2.3 in~\cite{ADV:2012}. Recall the notions that
appear in its formulation. Let $x_1,x_2,\dots,x_n,\dots$ be a sequence of letters. The sequence $\{Z_n\}_{n=1,2,\dots}$ of \emph{Zimin
words} is defined inductively by $Z_1=x_1$, $Z_{n+1}=Z_nx_{n+1}Z_n$. We say that an involutory word $v$ is an \emph{involutory isoterm for
a unary semigroup $\mathcal{S}$} if the only involutory word $v'$ such that $\mathcal{S}$ satisfies the \is\ identity $v=v'$ is the word
$v$ itself.

\begin{Thm}
\label{Theorem 2.2} Let $\mathcal{S}$ be a \fis. If all Zimin words are involutory isoterms for $\mathcal{S}$, then $\mathcal{S}$ is \infb.
\end{Thm}

\section{Main result and its applications}
\label{sec:main}

Recall that \sgps\ satisfying both $xy=yx$ and $x^2=x$ are called \emph{semilattices}. An \is\ $\mathcal{S}=\langle S,\cdot,{}^*\rangle$
whose reduct $\langle S,\cdot\rangle$ is a semilattice with 0 is said to be a \emph{twisted semilattice} if 0 is the only fixed point of
the involution $x\mapsto x^*$. This class of \iss\ was first considered in~\cite{Fajt}. It is easy to see that the minimum non-trivial
object in this class is the 3-element twisted semilattice $\TSL=\langle\{e,f,0\},\cdot,{}^*\rangle$ in which $e^2=e$, $f^2=f$ and all other
products are equal to $0$, while the unary operation is defined by $e^*=f$, $f^*=e$, and $0^*=0$.

If $ \mathcal{S}$ is an \is, we denote by $\var\mathcal{S}$ the variety generated by $\mathcal{S}$.

\begin{Thm}
\label{twisted} Let $\mathcal{S}=\langle S,\cdot,{}^*\rangle$ be a \fis\ such that $\TSL\in\var\mathcal{S}$. If the reduct $\langle
S,\cdot\rangle$ is \infb, then so is $\mathcal{S}$.
\end{Thm}

\begin{proof}
By Theorem~\ref{Theorem 2.2} we only have to show that $\mathcal{S}$ satisfies no non-trivial \is\ identity of the form $Z_n=z$. If $z$ is
a plain word, we can refer to~\cite[Proposition~7]{sapirburnside} according to which the INFB \sgp\ $\langle S,\cdot\rangle$ satisfies no
non-trivial plain \sgp\ identity of the form $Z_n=z$. Now suppose that $\mathcal{S}$ satisfies an identity $Z_n=z$ such that the involutory
word $z$ is not a plain word. This means that $z$ contains a starred letter. Since $\TSL\in\var\mathcal{S}$, the identity $Z_n=z$ holds in
$\TSL$. Substitute the element $e$ of $\TSL$ for all plain letters occurring in $Z_n$ and $z$. Since $e^2=e$, the value of the word $Z_n$
under this substitution equals $e$. On the other hand, since $z$ contains a starred letter, the value of $z$ is a product involving
$e^*=f$, and every such product is equal to either $f$ or 0. This is a contradiction.
\end{proof}

As for applications of Theorem~\ref{twisted}, we first give simplified and uniform proofs for two important results from \cite{ADV:2012}.
To start with, consider the \emph{twisted Brandt monoid} $\TB=\langle B_2^1,\cdot,{}^D\rangle$ arising when one endows the Brandt monoid
$\langle B_2^1,\cdot\rangle$ with the unary operation $A\mapsto A^D$ that fixes the matrices $\left(\begin{smallmatrix} 0 & 0\\ 0 &
0\end{smallmatrix}\right),\ \left(\begin{smallmatrix} 0 & 1\\ 0 & 0\end{smallmatrix}\right),\ \left(\begin{smallmatrix} 0 & 0\\ 1 &
0\end{smallmatrix}\right),\ \left(\begin{smallmatrix} 1 & 0\\ 0 & 1\end{smallmatrix}\right)$ and swaps each of the matrices
$\left(\begin{smallmatrix}1 & 0\\ 0 & 0\end{smallmatrix}\right),\ \left(\begin{smallmatrix} 0 & 0\\ 0 & 1\end{smallmatrix}\right)$ with the
other one. Observe that this unary operation is just the reflection with respect to the secondary diagonal (from the top right to the
bottom left corner). The reflection (called the \emph{skew transposition}) makes sense for every square matrix and is in fact an involution
of the \sgp\ $\langle\mathrm{M}_n(\mathcal{K}),\cdot\rangle$; this follows, for instance, from the observation that for every matrix
$A\in\mathrm{M}_n(\mathcal{K})$, one has $A^D=JA^TJ$, where $J$ is the $n\times n$-matrix with 1s on the secondary diagonal and 0s
elsewhere. It is a bit surprising that even though the \iss\ $\langle\mathrm{M}_n(\mathcal{K}),\cdot,{}^D\rangle$ and
$\langle\mathrm{M}_n(\mathcal{K}),\cdot,{}^T\rangle$ are isomorphic (under the map $A\mapsto JAJ$) and the set $B_2^1$ is closed under both
the usual transposition and the skew one, the involutary sub\sgps\ $\TB=\langle B_2^1,\cdot,{}^D\rangle$ and $\langle
B_2^1,\cdot,{}^T\rangle$ are essentially different. Indeed, the latter is not INFB (see Section~\ref{sec:preliminaries}) while the former
is, as the following corollary reveals.

\begin{Cor}[{\mdseries\cite[Corollary 2.7]{ADV:2012}}]
\label{twisted Brandt} The twisted Brandt monoid $\TB$ is inherently \nfb.
\end{Cor}

\begin{proof}
As already mentioned, the reduct $\langle B_2^1,\cdot\rangle$ of $\TB$ is INFB by~\cite[Corollary~6.1]{sapirburnside}. The matrices
$\left(\begin{smallmatrix}1 & 0\\ 0 & 0\end{smallmatrix}\right)$, $\left(\begin{smallmatrix}0 & 0\\ 0 & 1\end{smallmatrix}\right)$, and
$\left(\begin{smallmatrix}0 & 0\\ 0 & 0\end{smallmatrix}\right)$ form an involutory sub\sgp\ in $\TB$ and, obviously, this sub\sgp\ is
isomorphic to the 3-element twisted semilattice $\TSL$. Thus, Theorem~\ref{twisted} applies.
\end{proof}

Now consider the matrix \iss\ $\langle\mathrm{M}_n(\mathcal{K}),\cdot,{}^T\rangle$ where $\mathcal{K}$ is a finite field.

\begin{Cor}[{\mdseries\cite[Theorems 3.9 and 3.10]{ADV:2012}}]
\label{matrices} The \is\ $\langle\mathrm{M}_n(\mathcal{K}),\cdot,{}^T\rangle$, where $\mathcal{K}$ is a finite field, is \infb\ if $n\ge
3$ or if $n=2$ and the number of elements in $\mathcal{K}$ is not of the form $4k+3$.
\end{Cor}

\begin{proof}
The reduct $\langle\mathrm{M}_n(\mathcal{K}),\cdot\rangle$ is INFB for each $n\ge2$ and each finite field $\mathcal{K}$
by~\cite[Corollary~6.2]{sapirburnside}. To invoke Theorem~\ref{twisted}, it only remains to show that, under the condition of the
corollary, the 3-element twisted semilattice $\TSL$ belongs to the variety $\var\langle\mathrm{M}_n(\mathcal{K}),\cdot,{}^T\rangle$.

First let $n\ge3$. By the Chevalley-Warning theorem \cite[Corollary~2 in \S1.2]{Serre}, the field $\mathcal{K}$ contains some elements
$x,y$ satisfying $1+x^2+y^2=0$. Then the $n\times n$-matrices
$$e=\begin{pmatrix}
1 & 0 & 0 & \cdots & 0\\
x & 0 & 0 & \cdots & 0\\
y & 0 & 0 & \cdots & 0\\
\vdots & \vdots & \vdots &\ddots & \vdots\\
0 & 0 & 0 & \cdots & 0
\end{pmatrix},\ \ f=
\begin{pmatrix} 1 & x & y & \cdots & 0\\
0 & 0 & 0 & \cdots & 0\\
0 & 0 & 0 & \cdots & 0\\
\vdots & \vdots & \vdots &\ddots & \vdots\\
0 & 0 & 0 & \cdots & 0
\end{pmatrix},\ \text{ and }\
g=
\begin{pmatrix} 1 & x & y & \cdots & 0\\
x & x^2 & xy & \cdots & 0\\
y & xy & y^2 & \cdots & 0\\
\vdots & \vdots & \vdots &\ddots & \vdots\\
0 & 0 & 0 & \cdots & 0
\end{pmatrix}$$
satisfy
$$e^2=e,\ f^2=f,\ ef=g,\ fe=0,\ e^T=f,\ f^T=f,\ \text{ and } g^T=g.$$
Therefore the set $\{e,f,g,0\}$ forms an involutory sub\sgp\ in $\langle\mathrm{M}_n(\mathcal{K}),\cdot,{}^T\rangle$, the set $\{g,0\}$ is
an ideal of this sub\sgp and is closed under transposition. It remains to observe that the Rees quotient of the \is\
$\langle\{e,f,g,0\},\cdot,{}^T\rangle$ over the ideal $\{g,0\}$ is isomorphic to the 3-element twisted semilattice $\TSL$.

Now let $n=2$ and let the number of elements in $\mathcal{K}$ be not of the form $4k+3$. Then the field $\mathcal{K}$ contains an element
$x$ satisfying $1+x^2=0$, see, e.g., \cite[Theorem~3.75]{LidlNiederreiter}. Now the argument of the previous paragraph applies, with the
$2\times 2$-matrices
$$e'=\begin{pmatrix}
1 & 0 \\
x & 0
\end{pmatrix},\ \ f'=
\begin{pmatrix}
1 & x\\
0 & 0
\end{pmatrix},\ \text{ and }\
g'=
\begin{pmatrix}
1 & x \\
x & x^2
\end{pmatrix}$$
in the roles of $e,f$, and $g$, respectively.
\end{proof}

We could have continued along the same lines to show that in fact all examples of INFB \iss\ found in~\cite{Dolinka:2010,ADV:2012} can be
similarly deduced from Theorem~\ref{twisted}. However we think that Corollaries~\ref{twisted Brandt} and~\ref{matrices} are representative
enough. Now we present a new application.

Let $\mathcal{K}$ be a finite field and let $\mathrm{T}_n(\mathcal{K})$ stand for the set of all upper-triangular $n\times n$-matrices over
$\mathcal{K}$. The set $\mathrm{T}_n(\mathcal{K})$ forms an \is\ under the usual matrix multiplication and the skew transposition. The
following result classifies all INBF \iss\ of the form $\langle\mathrm{T}_n(\mathcal{K}),\cdot,{}^D\rangle$.

\begin{Thm}
\label{triangular} The \is\ $\langle\mathrm{T}_n(\mathcal{K}),\cdot,{}^D\rangle$, where $\mathcal{K}$ is a finite field, is \infb\ if and
only if $n\ge 4$ and $\mathcal{K}$ contains at least $3$ elements.
\end{Thm}

\begin{proof}
In~\cite{Goldberg&Volkov:2003} it is shown that the reduct $\langle\mathrm{T}_n(\mathcal{K}),\cdot\rangle$ is INFB if and only if $n\ge 4$
and $\mathcal{K}$ contains at least $3$ elements. Therefore, the `only if' part of our theorem follows from Lemma~\ref{easy} and the `if'
part will follow from Theorem~\ref{twisted} as soon as we will verify that $\TSL\in\var\langle\mathrm{T}_n(\mathcal{K}),\cdot,{}^D\rangle$.
Indeed, for every $n\ge2$ the matrix units $e_{11}$ and $e_{nn}$ belong to $\mathrm{T}_n(\mathcal{K})$ and satisfy
$$e_{11}^2=e_{11},\ e_{nn}^2=e_{nn},\ e_{11}e_{nn}=e_{nn}e_{11}=0,\ e_{11}^D=e_{nn},\ \text{ and } e_{nn}^D=e_{11}.$$
Hence the set $\{e_{11},e_{nn},0\}$ forms an involutory sub\sgp\ in $\langle\mathrm{T}_n(\mathcal{K}),\cdot,{}^D\rangle$ and this
involutory sub\sgp\ is isomorphic to the 3-element twisted semilattice $\TSL$.
\end{proof}

Observe that in~\cite{Goldberg&Volkov:2003} it is shown that for any $n$ and $\mathcal{K}$, the Brandt monoid $\langle B_2^1,\cdot\rangle$
does not belong to the \sgp\ variety generated by the \sgp\ $\langle\mathrm{T}_n(\mathcal{K}),\cdot\rangle$. Hence the twisted Brandt
monoid $\TB$ does not belong to the \is\ variety $\var\langle\mathrm{T}_n(\mathcal{K}),\cdot,{}^D\rangle$. Thus, Theorem~\ref{triangular}
provides a series of examples of INFB \iss\ whose varieties do not contain $\TB$. Such examples have not been known before.

\section{Regular semigroups}
\label{sec:regular}

In this section we address the question whether or not the presence of the 3-element twisted semilattice $\TSL$ in the variety generated by
a \fis\ $\mathcal{S}$ is necessary for the property of being INFB to be inherited by $\mathcal{S}$ from its \sgp\ reduct. So far we have no
answer to this question in its full generality; here we show that the answer is affirmative for the case where the reduct is a regular
\sgp.

We start with registering a property of Green's relations in \iss\ that do not have $\TSL$ among quotients of their involutory sub\sgps.

\begin{Lemma}
\label{D-structure} If $\TSL$ is not a quotient of an involutory sub\sgp\ $\mathcal{S}=\langle S,\cdot,{}^*\rangle$, then every idempotent
$e\in S$ satisfies $ee^*\Rc e\Lc e^*e\Rc e^*\Lc ee^*$.
\end{Lemma}

\begin{proof}
Clearly, we may assume that $\mathcal{S}$ is generated by $e$ as a unary \sgp. Let $I=See^*S\cup Se^*eS$, then $I$ is a $*$-closed ideal in
$\mathcal{S}$. If $e\notin I$, then also $e^*\notin I$ and the Rees quotient $\mathcal{S}/I$ is isomorphic to $\TSL$, a contradiction.
Thus, $e\in I$, that is $e=w$ where $w$ is an alternating product of factors $e$ and $e^*$. Multiplying the equality $e=w$ through by $e$
on the left and on the right, we obtain an equality of the form $e=ee^*\cdots e^*e$. Hence $ee^*\Rc e\Lc e^*e$. Using the fact that the
involution is an anti-isomorphism of $\langle S,\cdot\rangle$ and the equalities $(ee^*)^*=ee^*$, $(e^*e)^*=e^*e$, we also obtain $ee^*\Lc
e^*\Rc e^*e$.
\end{proof}

Recall the classic result by Miller and Clifford that will be invoked a few times.

\begin{Lemma}[{\mdseries\cite[Proposition~2.3.7]{how}}]
\label{Miller&Clifford} Let $a,b$ be elements of a $\mathscr{D}$-class of an arbitrary \sgp. Then $ab\in R_a\cap L_b$ if and only if
$L_a\cap R_b$ contains an idempotent.
\end{Lemma}

We will also need the following result from~\cite{ADV:2012}.

\begin{Prop}[{\mdseries\cite[Proposition~2.9]{ADV:2012}}]
\label{NINFB} Let $\mathcal{S}=\langle S,\cdot,{}^*\rangle$ be a \fis. If there exists an involutory word $\om(x)$ in one variable $x$ such
that $\mathcal{S}$ satisfies the identity $x=x\om(x)x$, then $\mathcal{S}$ is not \infb.
\end{Prop}

Recall that an element $x$ of a \sgp\ $\langle S,\cdot\rangle$ is said to be \emph{regular} if $x=xyx$ for some $y\in S$. A [unary] \sgp\
is \emph{regular} if all of its elements are regular.

\begin{Thm}
\label{regular} Let $\mathcal{S}=\langle S,\cdot,{}^*\rangle$ be a finite regular \is. If $\TSL\notin\var\mathcal{S}$, then $\mathcal{S}$
is not \infb.
\end{Thm}

\begin{proof}
Take an arbitrary element $x\in S$ and fix an element $y\in S$ such that $x=xyx$. Then $e=xy$ and $f=yx$ are idempotents and $e\Rc x\Lc f$.
Since the involution is an anti-isomorphism of $\langle S,\cdot\rangle$, we also have $e^*\Lc x^*\Rc f^*$. By Lemma~\ref{D-structure},
$ee^*\Rc e\Lc e^*e\Rc e^*\Lc ee^*$ and $ff^*\Rc f\Lc f^*f\Rc f^*\Lc ff^*$.  All listed relations are graphically represented in
Fig.~\ref{fig:D-class} that shows an appropriate fragment of the eggbox picture for the $\mathscr{D}$-class of $x$ and $x^*$.
\begin{figure}[th]
\begin{center}
{\large \begin{tabular}{|c|c|c|c|}
\hline $x$\rule[-5pt]{0pt}{16pt} & $e$ & & $ee^*$\\
\hline \rule[-5pt]{0pt}{16pt}& $e^*e$  & & $e^*$\\
\hline $f$\rule[-5pt]{0pt}{16pt} & \phantom{$f^*f$} & $ff^*$ &  \phantom{$f^*f$}\\
\hline $f^*f$\rule[-5pt]{0pt}{16pt} & & $f^*$ & $x^*$ \\
\hline
\end{tabular}}
\caption{A fragment of the eggbox picture for the $\mathscr{D}$-class of the elements $x$ and $x^*$}\label{fig:D-class}
\end{center}
\end{figure}
\end{proof}

By the `only if' part of Lemma~\ref{Miller&Clifford}, the fact that the product $e^*e$ belongs to $L_e\cap R_{e^*}$ implies that the
$\mathscr{H}$-class $H_{ee^*}=R_e\cap L_{e^*}$ contains an idempotent $g$ and hence, by Green's theorem~\cite[Theorem~2.2.5]{how}, this
class is a subgroup of $\langle S,\cdot\rangle$ having $g$ as its identity element. Observe that $R_e=R_x$ and $L_{e^*}=L_{x^*}$ whence
$H_{ee^*}=R_x\cap L_{x^*}$. Also observe that $gx=x$ since $g$ is an idempotent and $g\Rc x$.

Similarly, $ff^*\in R_f\cap L_{f^*}$ implies that $H_{f^*f}=R_{f*}\cap L_f$ contains an idempotent. However, $R_{f^*}=R_{x^*}$ and
$L_f=L_x$. Now, by the `if' part of Lemma~\ref{Miller&Clifford}, the fact that $L_x\cap R_{x^*}=H_{f^*f}$ contains an idempotent implies
that the product $xx^*$ lies in $R_x\cap L_{x^*}=H_{ee^*}$. Let $n$ be the least common multiple of exponents of the subgroups of $\langle
S,\cdot\rangle$. We then have $(xx^*)^n=g$ whence $(xx^*)^nx=x$.

Since $x\in S$ is arbitrary, we see that the \is\ $\mathcal{S}$ satisfies the identity $(xx^*)^nx=x$. Now Proposition~\ref{NINFB} applies
(with $\om(x)=x^*(xx^*)^{n-1}$).

Now we may deduce various characterizations of regular INFB \iss.
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{Cor}
\label{characterization} Let $\mathcal{S}=\langle S,\cdot,{}^*\rangle$ be a finite regular \is. Then the following are equivalent:
\begin{enumerate}
\item $\mathcal{S}$ is \infb;
\item the reduct $\langle S,\cdot\rangle$ is \infb\ and the variety $\var\mathcal{S}$ contains the $3$-element twisted
semilattice $\TSL$;
\item all Zimin words are involutory isoterms for $\mathcal{S}$.
\end{enumerate}
\end{Cor}

\begin{proof}
The implication (i) $\to$ (ii) follows from Lemma~\ref{easy} and Theorem~\ref{regular}. The implication (ii) $\to$ (iii) has been verified
in the course of the proof of Theorem~\ref{twisted}. The implication (iii) $\to$ (i) is Theorem~\ref{Theorem 2.2}.
\end{proof}

Finally, we observe that the condition (ii) in Corollary~\ref{characterization} is algorithmically verifiable. Indeed, given a finite
regular \is\ $\mathcal{S}$, we can check whether or not its reduct is INFB by using Sapir's algorithm from~\cite{Sapir:1987}. In order to
verify whether or not the variety $\var\mathcal{S}$ contains the $3$-element twisted semilattice $\TSL$, we can use a general algorithm by
Kalicki~\cite{Kalicki:1952}. This algorithm (being in fact an easy consequence of the standard proof of the HSP-theorem) decides, when
given two finite algebras $\mathcal{A}$ and $\mathcal{B}$ of the same similarity type, whether or not $\mathcal{A}$ lies in the variety
generated by $\mathcal{B}$. We have thus established the following fact.

\begin{Cor}
There exists an algorithm which decides, when given a  finite regular \is\ $\mathcal{S}$, whether or not $\mathcal{S}$ is \infb.
\end{Cor}


\begin{thebibliography}{99}
\bibitem{ADV:2012}
Auinger, K., Dolinka I., Volkov, M.V.: Matrix identities involving multiplication and transposition. J. Europ. Math. Soc. \textbf{14}(3),
937--969 (2012)

\bibitem{BuSa81}
Burris, S., Sankappanavar, H.P.: A Course in Universal Algebra. Springer, Berlin (1981)

\bibitem{Dolinka:2010}
Dolinka, I.: On identities of finite involution semigroups. Semigroup Forum \textbf{80}, 105--120 (2010)

\bibitem{Fajt}
Fajtlowicz, S.: Equationally complete semigroups with involution. Algebra Universalis \textbf{1}, 355--358 (1972)

\bibitem{Goldberg&Volkov:2003}
Goldberg, I.A., Volkov, M.V.: Identities of semigroups of triangular matrices over a finite field. Mat. Zametki. \textbf{73}(4), 502--510
(2003), in Russian; Engl.\ translation Math.\ Notes \textbf{73}(4), 474--481 (2003)

\bibitem{how}
Howie, J.M.: Fundamentals of Semigroup Theory. London Mathematical Society Monographs \textbf{12}, Clarendon Press, Oxford (1995)

\bibitem{Jackson&Volkov:2010}
Jackson, M., Volkov, M.V.: The algebra of adjacency patterns: Rees matrix semigroups with reversion. In: Fields of Logic and Computation.
Lecture Notes in Compiter Science, vol. 6300, pp. 414--443. Springer, Berlin (2010)

\bibitem{Kalicki:1952}
Kalicki, J.: On comparison of finite algebras. Proc.\ Amer.\ Math.\ Soc. \textbf{3}(1), 36--40 (1952)

\bibitem{Lawrence&Willard:1998}
Lawrence, J.,  Willard, R.: On finitely based groups and nonfinitely based quasivarieties. J. Algebra \textbf{203}, 1--11 (2008)

\bibitem{LidlNiederreiter}
Lidl, R., Niederreiter, H.: Finite Fields. Addison-Wesley, Cambridge (1997)

\bibitem{sapirburnside}
Sapir, M. V.: Problems of Burnside type and the finite basis property in varieties of semigroups. Izv.\ Akad.\ Nauk SSSR, Ser.\ Mat.
\textbf{51}, 319--340 (1987), in Russian; Engl.\ translation Math.\ USSR-Izv. \textbf{30}, 295--314 (1987)

\bibitem{Sapir:1987}
Sapir, M.V.: Inherently nonfinitely based finite semigroups. Mat. Sb. \textbf{133}(2), 154--166 (1987), in Russian; Engl.\ translation
Math. USSR--Sb. \textbf{61}, 155--166 (1988)

\bibitem{Sapir:1993}
Sapir, M. V.: Identities of finite inverse semigroups. Internat.\ J. Algebra Comput. \textbf{3}, 115--124 (1993)

\bibitem{Serre}
Serre, J.-P.: Cours d'Arithmetique. Presses Universitaires de France, Paris (1980)

\bibitem{Volkov:2001}
Volkov, M.V.: The finite basis problem for finite semigroups. Sci. Math. Jpn. \textbf{53}(1), 171--199 (2001)
\end{thebibliography}

\end{document}
